---
title: "Project 2 Group 4"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Project 2 Group 4
## von Anna, Ann-Sophie und Jana

new dataframe with only erlotinib columns: 249 - 307
```{r }
erlotinib_treated <- NCI_TPW_gep_treated[, 249:307]
erlotinib_untreated <- NCI_TPW_gep_untreated[, 249:307]
```


save as dataframe
```{r }
erlotinib_treated <- as.data.frame(erlotinib_treated)
erlotinib_untreated <- as.data.frame(erlotinib_untreated)
```

combine data of treated and untreated in one 
```{r }
erlotinib_all <- cbind (erlotinib_treated, erlotinib_untreated)
```

Principal component analysis
```{r }
pca <- prcomp(erlotinib_all)
plot(pca, type = "l")
plot(pca$rotation[,1], pca$rotation[,2])

color <- ifelse(colnames(erlotinib_all) == colnames(erlotinib_treated), "forestgreen", "firebrick")
plot(pca$rotation[,1], pca$rotation[,2], col = color)
```

Z-transormation
```{r }
median_gene <- apply (erlotinib_all, 1, function (x){
+ median(x) })
median(median_gene)
sd_gene <- apply(erlotinib_all, 1, function (x){
 + sd(x) })
sd (sd_gene)
erlotinib_all_z_transformed <- (erlotinib_all-median(median_gene)) / sd(sd_gene)
```

erlotinib_normalized_z_transformed: treated-untreated + z-Transformation + PCA 
```{r }
erlotinib_normalized <- erlotinib_treated - erlotinib_untreated

median_gene <- apply (erlotinib_normalized, 1, function (x){
+ median(x) })
sd_gene <- apply(erlotinib_normalized, 1, function (x){
+ sd(x) })
erlotinib_normalized_z_transformed <- (erlotinib_normalized-median(median_gene)) / sd(sd_gene)

pca <- prcomp(erlotinib_normalized_z_transformed)
plot(pca$rotation[,1], pca$rotation[,2])
```

select cell lines with highest variance
```{r}
var_cell_line <- apply(erlotinib_normalized_z_transformed, 2, function (x) {
+ var(x)})
cell_line_var_greater_75quantile <- var_cell_line [which (var_cell_line > quantile(var_cell_line,0.75))]
cell_line_var_decreasing_top15 <- sort(cell_line_var_greater_75quantile, decreasing = TRUE)

table_cell_lines_var_top15 <- cbind(names(cell_line_var_decreasing_top15), cell_line_var_decreasing_top15)
rownames(table_cell_lines_var_top15) <- c(1:15)
colnames(table_cell_lines_var_top15) <- c("cell line", "variance")
```

PCA visualization with factoextra package: 
```{r}
install.packages("factoextra")
library(factoextra)
fviz_eig(pca) #shows percentage explained by each PC
fviz_pca_ind(pca, col.ind = "cos2", gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"))
```

PCA with transformed matrix (each dot represents a gene):
```{r}
transformed_erlotinib_normalized_z_transformed <- t(erlotinib_normalized_z_transformed)
pca <- prcomp(transformed_erlotinib_normalized_z_transformed)
plot(pca$rotation[,1], pca$rotation[,2])
text(pca$rotation, labels = rownames(erlotinib_normalized_z_transformed), cex = 0.4, pos = 3)
```

PCA visualization with factoextra package: 
```{r}
install.packages("factoextra")
library(factoextra)
fviz_eig(pca) #shows percentage explained by each PC
fviz_pca_ind(pca, col.ind = "cos2", gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"))
```


