---
title: "specific analysis - 2. Milestone"
output: github_document

## von Anna, Ann-Sophie und Jana

```{r, include=FALSE}
#Load required packages
library(knitr)
```

```{r setup, include=FALSE} 
#code and output not included in final file
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache = TRUE)
```


## Load data

```{r }
NCI_TPW_gep_treated <- readRDS(url("https://ndownloader.figshare.com/files/14720180?private_link=db1411debcfbe2618d2f"))
NCI_TPW_gep_untreated <- readRDS(url("https://ndownloader.figshare.com/files/14720183?private_link=db1411debcfbe2618d2f"))
NCI_TPW_metadata <- read.delim("https://ndownloader.figshare.com/files/14720186?private_link=db1411debcfbe2618d2f")
NegLogGI50 <- readRDS(url("https://ndownloader.figshare.com/files/14720210?private_link=074e0120fe5e682f3d14"))
CCLE_basalexpression <- readRDS(url("https://ndownloader.figshare.com/files/14770127?private_link=fc0c71246dc192982a3c"))
CCLE_copynumber <- readRDS(url("https://ndownloader.figshare.com/files/14770130?private_link=fc0c71246dc192982a3c"))
CCLE_mutations <- readRDS(url("https://ndownloader.figshare.com/files/14770133?private_link=fc0c71246dc192982a3c"))
cellline_annotation <-read.delim("https://ndownloader.figshare.com/files/14768981?private_link=efb6a529eaf2d4dc6432")
drug_annotation <- read.delim("https://ndownloader.figshare.com/files/14768984?private_link=efb6a529eaf2d4dc6432")
```

#new matrix only with samples/columns treated with erlotinib  (e=erlotinib)
```{r}
e_treated <- NCI_TPW_gep_treated[,grep ("erlotinib", colnames(NCI_TPW_gep_treated))]
e_untreated <- NCI_TPW_gep_untreated[,grep ("erlotinib", colnames(NCI_TPW_gep_untreated))]
e_foldchange <- e_treated - e_untreated
```

#colnames of e_foldchange with cellline instead of complete sample name
  cellline <- sapply(colnames(e_foldchange), function(x){
  colnames(annotation_sorted_by_cell_lines)[which(x == annotation_sorted_by_cell_lines, arr.ind = TRUE)[2]]
})
colnames(e_foldchange) <- cellline

#e_foldchange_normalized: z-Transformation to get mean=0 and sd=1
```{r}
e_foldchange_normalized <- apply(e_foldchange, 2, function(x){
  (x - mean(x)) / sd(x)
})
```


# table of 15 cell lines with highest variance to show most regulated cell lines
#select 15 cell lines with highest variance (greater than 75% quantile, sorted by decreasing value)
```{r}
var_cell_line <- apply(e_foldchange, 2, var)
cell_line_var_greater_75quantile <- sort(var_cell_line [which (var_cell_line > quantile(var_cell_line,0.75))], decreasing = TRUE)
rm(var_cell_line)
```

#create a table containing the name and the variance of the 15 cell lines with highest variance

```{r}
table_cell_lines_var_top15 <- cbind(names(cell_line_var_greater_75quantile), cell_line_var_greater_75quantile)
rownames(table_cell_lines_var_top15) <- c(1:nrow(table_cell_lines_var_top15))
colnames(table_cell_lines_var_top15) <- c("cell line", "variance")
```


#PCA to show most regulated cell lines
#PCA with transformed matrix (each point represents a sample):

```{r}
e_foldchange_matrix_transformed <- t(e_foldchange_normalized)
pca <- prcomp(e_foldchange_matrix_transformed)
plot(pca$rotation[,1], pca$rotation[,2])
text(pca$rotation, labels = rownames(e_foldchange_normalized), cex = 0.4, pos = 3)
```

#PCA visualization with factoextra package (according to www.sthda.com/english/articles/31-principal-component-methods-in-r-practical-guide/118-principal-component-analysis-in-t-prcomp-vs-princomp/) : 
#plot of cell lines with gradient colors by their quality of representation
 #cos2 values correspond to their quality of representation
install.packages("factoextra")
library(factoextra)
fviz_pca_ind(pca, col.ind = "cos2", gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"))

#bar plot showing the percentage explained by each PC
#plot of genes showing their contribution to the two PCs 
#get contribution of genes to the PCs
fviz_eig(pca)
fviz_pca_var(pca, col.var ="contrib", gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"))
results.genes <- get_pca_var(pca)
genes_pca_highest_contribution <- results.genes$contrib '



#Volcano plot
#mean of gene expression of each gene over all cell lines

```{r}
e_foldchange_mean_over_cell_lines <- rowMeans(e_foldchange) #equal to e_treated_mean_over_cell_lines - e_untreated_mean_over_cell_line
```

#determine the p-value for a paired two-sample t-test 
```{r}
p_values <- sapply(rownames(e_treated), function(x) {
  t.test(e_treated[x,], e_untreated[x,],paired= T)$p.value}) # perform t-test and save p-values of each gene in p_vales-vector
FDR_values <- p.adjust(p_values, method = "BH", n = length(p_values))#calculate FDR with benjamini-hochberg (BH)
```


#table of results 
#coloring with package enhanced volcano
#install package EnhancedVolcano (needs ggplot2, ggrepel)

```{r}
statistics_values <- cbind(e_foldchange_mean_over_cell_lines, FDR_values)
if (!requireNamespace('BiocManager', quietly = TRUE))
  install.packages('BiocManager')
BiocManager::install('EnhancedVolcano')
library(EnhancedVolcano)

EnhancedVolcano(statistics_values, 
                lab = rownames(statistics_values),
                x = "e_foldchange_mean_over_cell_lines", #colname of FC values in this table (statistics_values)
                y = "FDR_values", #colname of FDR (statistics_values)
                title = "Volcano plot of all genes",
                pCutoff = 10e-15, #threshold for coloring significant ones
                FCcutoff = 1, #threshold for coloring high FC
                transcriptPointSize = 2,
                transcriptLabSize = 4.0)
```



#further questions: what "do" the red genes? are they involved in certain pathways?
#maybe we could color the genes in the volcano plot according to pathways (after milestone 4)

#save the "red" genes seen in the volcano plot in a vector for further analysis
```{r}
biomarkers <- rownames(statistics_values)[which(abs(statistics_values[, 1]) > 1 
                                                   & statistics_values[, 2] < 10e-15)]
```


#Density plot with these genes (untreated vs. treated)
```{r}
plot(density(e_treated[biomarkers, ]), "Density plot of gene expression", col = "red")
lines(density(e_untreated[biomarkers, ]), col = "black")
legend("topright", legend = c("untreated", "treated"), col = c("black", "red"), pch = 15)
```


#MA-Plot
##install package and load ggplot2 and ggrepel
```{r}
install.packages("ggplot2")
library(ggplot2)
library(ggrepel)
```

#create matrices with the variables M and A of a MA-plot
```{r}
M <- e_foldchange # M= log2(treated) - log2 (untreated)
A <- 1/2*(e_treated+ e_untreated) # average log2-expression value A = 1/2 (log2(treated)+log2(untreated))
MA <- cbind("M"= rowMeans(M), "A" = rowMeans(A), FDR_values)
MA <- as.data.frame(MA)
MA$Significant <- ifelse(MA$FDR_values<0.05, "FDR < 0.05", "Not Sig")
```

#matrix with important genes of MA plot 
```{r}
MA_labeled <- MA[which(MA[ , "M"] > 1.5 | MA[,"M"] > 0.95 & MA[,"A"] > 10) , ]
```

#MA plot labeled with important genes of MA plot
```{r}
ggplot(data=MA)+ 
  aes(x=A, y=M, color= Significant)+
  geom_point()+
  xlab("mean expression")+
  ylab("log fold change")+
  geom_text(data=MA_labeled, aes(A, M, label=rownames(MA_labeled)))
```

#label biomarkers, obtained through volcano plot, in MA plot
```{r}
MA_biomarkers <- sapply(biomarkers, function(x){
  MA[x, ]
})
MA_biomarkers <- t(MA_biomarkers)
MA_biomarkers <- as.data.frame(MA_biomarkers)
```

#MA plot labeled with biomarker genes of volcano plot
```{r}
ggplot(data=MA)+ 
  aes(x=A, y=M, color= Significant)+
  geom_point()+
  xlab("mean expression")+
  ylab("log fold change")+
  geom_text(data=MA_biomarkers, aes(A, M, label=rownames(MA_biomarkers)))
```

# boxplot of foldchange of biomarkers 
## create a matrix foldchange_biomarkers, with the foldchange only of the biomarkers
```{r}
foldchange_biomarkers <- sapply(biomarkers, function(x){
  e_foldchange[x, ]
})
boxplot(foldchange_biomarkers, ylab= "foldchange", 
        main= "boxplot of foldchange of the biomarkers", las=2)
```


# boxplot of gene expression treated vs. untreated of biomarkers 
## create a matrix e_treated_biomarkers/ e_untreated_biomarkers, with the gene expression only of the biomarkers
```{r}
e_treated_biomarkers <- sapply(biomarkers, function(x){
  e_treated[x, ]
})
e_untreated_biomarkers <- sapply(biomarkers, function(x){
  e_untreated[x, ]
})
```

# #create a matrix, which contains gene expression of treated and untreated and sort it after colnames
```{r}
e_treated_untreated_biomarkers <- cbind (e_treated_biomarkers, e_untreated_biomarkers)
e_treated_untreated_biomarkers <- e_treated_untreated_biomarkers[,order(colnames(e_treated_untreated_biomarkers))]
```

## boxplot, where treated and untreated are right next to each other 
```{r}
boxplot(e_treated_untreated_biomarkers, ylab= "gene expression (log2)", 
        main= "boxplot of gene expression of the biomarkers", las=2)
```



